---
execute: 
  echo: true
  cache: true
---

# Collaboration {#sec-collaboration}


```{r}
#| label: setup-options
#| child: "_common.qmd"
```

::: {.callout-caution title="Caution"}
Under development.
:::

<!-- Other protocols/ examples:

[Project TIER](https://www.projecttier.org/)

-->

<!-- [ ] Take a look at @Rodrigues2023 for a good overview of the research process. This includes using *renv* (https://rstudio.r-universe.dev/renv#) for package versions, *targets* (https://ropensci.r-universe.dev/targets#) for workflow pipelines, and Rocker project (https://rocker-project.org/) for containerization. -->

> The reproducibility of studies and the ability to follow up on the work of others is key for innovation in science and engineering. 
>
> --- Leland Wilkinson

::: callout-note
## Keys

- ...
:::

```{r}
#| label: reporting-data-packages
#| echo: false
#| message: false
# Packages

pacman::p_load(tidyverse)
```

Whether for other researchers or for your future self, creating research that is well-documented and reproducible is a fundamental part of conducting modern scientific inquiry. In this chapter we will emphasize the importance of this endeavor and outline strategies for ensuring your research project is reproducible. This will include directory and file structure, key documentation files as well as how to effectively use existing software resources and frameworks for publishing your research (either for private use, journal requirements, or general public consumption) on popular repositories such as GitHub and Open Science Framework (OSF).

::: callout-tip
## Swirl

**What**: [Compiling Research Projects](https://github.com/qtalr/lessons)\
**How**: In the R Console pane load `swirl`, run `swirl()`, and follow prompts to select the lesson.\
**Why**: To ...
:::

Outline...

This is what we will cover in this chapter: ...




## Questions {.unnumbered}

::: callout-note
## Conceptual questions

1. ...
2. ...

:::

::: callout-note
## Technical exercises

1. ...
2. ...
:::

