---
execute: 
  echo: true
---

# Reports {#sec-reports}

```{r}
#| label: setup-options
#| child: "_common.qmd"
#| cache: false
```

::: {.callout-caution title="Caution"}
Under development.
:::

<!--

Content:

Exercises:

- [ ] add concept questions to Activities
- [ ] add exercises to Activities
- [ ] add thought questions/ case studies to prose sections

Formatting:


-->

> ..quote..
>
> --- Author

::: callout-note
## Keys

- ...
:::

```{r}
#| label: reporting-data-packages
#| echo: false
#| message: false
# Packages

pacman::p_load(tidyverse)
```

In this chapter, we first examine the role of research presentations in academic and professional settings, highlighting the benefits of presenting one's work as a means to develop and refine ideas and conclusions. We then discuss the purpose of a research document, focusing on how its structure effectively conveys the project's rationale, goals, procedures, results, and findings. Additionally, we explore the integration of citations, references, figures, and tables in research documents using Quarto. Lastly, we investigate the application of field-specific and publishing house formats to a variety of document formats, such as Word, PDF, HTML, and ePub.

::: callout-tip
## Swirl

**What**: [Rendering Quarto](https://github.com/qtalr/lessons)\
**How**: In the R Console pane load `swirl`, run `swirl()`, and follow prompts to select the lesson.\
**Why**: To ...
:::

1. Introduction
   - Importance of clear and effective communication in research
       - Enhances understanding of the research
       - Facilitates collaboration and interdisciplinary work
   - Overview of the chapter content
       - Research presentations
       - Research document structure
       - Citations and references
       - Figures and tables
       - Field-specific and publishing house formats

2. Research Presentations
   - Importance of research presentations in academia and professional contexts
       - Sharing research findings
       - Building a professional reputation
   - Benefits of presenting research work
       - Developing and refining ideas and conclusions through feedback and discussion
       - Receiving feedback and suggestions from peers and experts
         - Conducting peer review can help ensure that the research is conducted appropriately and transparently. Peer review can also identify potential errors or biases in the research design or analysis, and can provide suggestions for improving the study's reproducibility
       - Enhancing communication and presentation skills
       - Building professional network and fostering collaboration
   - Key elements of an effective research presentation
       - Structure and organization: logical flow, clear sections, concise points
       - Visual design and aesthetics: use of color, fonts, visuals, and layout
       - Delivery and engagement: pacing, tone, body language, and audience interaction
   - Using R for creating research presentations
       - RMarkdown and its integration with presentation formats (e.g., xaringan, slidify)
       - Incorporating data visualization, tables, and other R-generated content seamlessly
   - Tips for delivering engaging and memorable presentations
       - Practice and preparation
       - Storytelling and relatability
       - Addressing diverse audience backgrounds

3. Research Documents
   - Purpose of a research document
       - Communicate research methodology and findings
       - Contribute to the body of knowledge
   - Benefits of writing a research document
       - Clarifying and organizing thoughts and ideas
       - Providing a comprehensive record of research work
       - Enhancing writing and argumentation skills
       - Contributing to the body of knowledge in a field
       - Establishing professional credibility and visibility
   - Components of a research document
       - Title: concise and informative
       - Abstract: brief summary of the research
       - Introduction: background, research question, and objectives
       - Methodology: description of the data, tools, and analysis techniques
       - Results: presentation of the findings
       - Discussion: interpretation and implications of the results
       - Conclusion: summary and future directions
       - References: list of cited sources
   - Tips for organizing and structuring research documents
       - Outlining the main sections
       - Maintaining logical flow and consistency
       - Using clear and concise language

4. Citations and References
   - Importance of proper citation and referencing
       - Acknowledge the work of others
       - Demonstrate the foundation of your research
   - Different citation styles (APA, MLA, Chicago, etc.)
       - Overview of common styles
       - Choosing the appropriate style for your discipline
   - Using R and Quarto for managing citations and references
       - Integration with reference managers (e.g., Zotero, Mendeley)
       - Automating citation formatting
   - Integration of citations and references in research documents
       - In-text citations
       - Reference list or bibliography

5. Figures and Tables
   - Importance of visual representations in research reporting
       - Aid in understanding complex data
       - Enhance the readability of the document
   - Creating and customizing figures and tables using R and Quarto
       - ggplot2 for creating visualizations
       - kable and gt for generating tables
   - Guidelines for labeling, formatting, and presenting figures and tables
       - Clear and informative titles and labels
       - Consistent formatting
       - Appropriate use of color, fonts, and layout
       - Proper referencing and integration in the text
       - Accessibility considerations for diverse readers

6. Field-specific and Publishing House Formats
   - Importance of adhering to field-specific and publisher guidelines
       - Consistency and professionalism
       - Meeting submission requirements for journals, conferences, and other venues
   - Overview of common format requirements
       - Manuscript formatting: margins, line spacing, headings, etc.
       - Citation and reference style
       - Figure and table formatting
   - Using R and Quarto to apply formatting guidelines
       - Customizing document templates
       - Applying and managing style files
   - Exporting research documents to various formats
       - Word: .docx files for collaboration and editing
       - PDF: .pdf files for sharing and printing
       - HTML: web-based documents for online publication
       - ePub: e-book format for digital reading

7. Conclusion
   - Recap of the main topics covered in the chapter
   - Importance of clear and effective communication in research reporting
   - Encouragement to practice and refine reporting skills through various research projects




## Questions {.unnumbered}

::: callout-note
## Conceptual questions

1. ...
2. ...

:::

::: callout-note
## Technical exercises

1. ...
2. ...
:::