---
execute: 
  echo: true
---

# Reporting {#sec-reporting}

```{r}
#| label: setup-options
#| child: "_common.qmd"
#| cache: false
```

<!--

Content:

Exercises:

- [ ] add concept questions to Activities
- [ ] add exercises to Activities
- [ ] add thought questions/ case studies to prose sections

Formatting:


-->

> ..quote..
>
> --- Author

::: callout-note
## Keys

- ...
:::

```{r}
#| label: reporting-data-packages
#| echo: false
#| message: false
# Packages

pacman::p_load(tidyverse)
```

First we will discuss the purpose of a research document and how the structure of this document supports and clearly documents the project’s rationale, goals, procedure, results, and findings. Then I will turn to incorporating citations and references and figures and tables in Quarto. Finally, we will explore how to apply field-specific and publishing house formats to various document formats including Word, PDF, HTML, and ePub.

::: callout-tip
## Swirl

**What**: [Rendering Quarto](https://github.com/lin380/swirl)\
**How**: In the R Console pane load `swirl`, run `swirl()`, and follow prompts to select the lesson.\
**Why**: To ...
:::

Outline...




## Questions {.unnumbered}

::: callout-note
## Conceptual questions

1. ...
2. ...

:::

::: callout-note
## Technical exercises

1. ...
2. ...
:::